---
title: "Chapter 12 - Factors"
author: "Daniel Czarnievicz"
date: "5/20/2019"
output: 
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    toc_float: 
      smooth_scroll: true
    theme: united
    highlight: tango
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "")
library(tidyverse)
library(magrittr)
library(stringr)
library(forcats)
```

# Creating factors

## Section notes

```{r}
# To creat a factor, start by creating a vector of the valid levels
month_levels <- c( "Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
x1 <- c("Dec", "Apr", "Jan", "Mar")
y1 <- factor(x1, levels = month_levels)

# Values not in levels will be coerced to NA
x2 <- c("Dec", "Apr", "Jam", "Mar")
y2 <- factor(x2, levels = month_levels)

# Unlike character vectors, factors will sort based on levels
sort(x1)
sort(y1)

# If you want error reports, parse the factors
parse_factor(x2, levels = month_levels)

# If levels are not set explicitly, they will be done alphabetically
factor(x1)

# To set them based on first appearence use unique or fct_indorder
f1 <- factor(x1, levels = unique(x1))
f2 <- x1 %>% factor() %>% fct_inorder()

# To access the levels, use levels()

```

# GSS

## Section notes

```{r}
# Visualizing the factors
count(gss_cat, race)
ggplot(gss_cat, aes(race)) +
      geom_bar()
ggplot(gss_cat, aes(race)) +
      geom_bar() +
      scale_x_discrete(drop = FALSE)

```

## Exercise set 1 (page 227)

### 1 - Explore `rincome`

```{r}
ggplot(gss_cat) +
      geom_bar(aes(rincome))
```

### 2 - Most common `relig` and `partyid`

```{r}
count(gss_cat, relig) %>% arrange(desc(n))
count(gss_cat, partyid) %>% arrange(desc(n))
```

### 3 - Religion for `denom`

```{r}
gss_cat %>% 
      group_by(relig, denom) %>% 
      tally() %>% 
      arrange(desc(n))
ggplot(gss_cat) +
      geom_count(aes(x = relig, y = denom)) +
      theme(axis.text.x = element_text(angle = 60, hjust = 1))

```

# Modifying factor order

## Section notes

```{r}
# Changing the order of factor levels for visualization
relig <- gss_cat %>%
      group_by(relig) %>%
      summarize(age = mean(age, na.rm = TRUE),
                tvhours = mean(tvhours, na.rm = TRUE),
                n = n())
ggplot(relig, aes(tvhours, relig)) + geom_point()
```

```{r}
# fct_reorder takes three arguments:
#     .f the factor to reorder
#     .x numeric vector used to reorder
#     fun a function that's used if there are multiple values of .x for each value of .f (default is median)
ggplot(relig) +
      geom_point(aes(tvhours, fct_reorder(.f = relig, .x = tvhours)))
```

```{r}
# for more complicated transformations, Hadely recomends taking reorder out of aes and into mutate
relig %>%
      mutate(relig = fct_reorder(relig, tvhours)) %>%
      ggplot(aes(tvhours, relig)) +
      geom_point()
```

```{r}
# Reordering factors dosen't always make sense
rincome <- gss_cat %>%
      group_by(rincome) %>%
      summarize(age = mean(age, na.rm = TRUE),
                tvhours = mean(tvhours, na.rm = TRUE),
                n = n())
ggplot(rincome, aes(age, fct_reorder(rincome, age))) + 
      geom_point()
```

```{r}
# To move certain factor levels to the front of the line, use fct_relevel
ggplot(rincome, aes(age, fct_relevel(rincome, "Not applicable"))) +
      geom_point()
```

```{r, out.width="50%"}
# reorder2 reorders the factor by the y values associated with the largest x values
by_age <- gss_cat %>%
      filter(!is.na(age)) %>%
      count(age, marital) %>%
      group_by(age) %>%
      mutate(prop = n / sum(n))
ggplot(by_age, aes(age, prop, color = marital)) +
      geom_line(na.rm = TRUE)
ggplot(by_age, aes(age, prop, color = fct_reorder2(marital, age, prop))) +
      geom_line() +
      labs(color = "marital")
```

```{r}
# for bar plots, use fct_infreq with fct_rev
#     fct_infreq takes a factor f and reorders based on it's frequency
#     fct_rev takes a factor and reverses its order
gss_cat %>%
      mutate(marital = marital %>% fct_infreq() %>% fct_rev()) %>%
      ggplot(aes(marital)) +
      geom_bar()
```

## Exercise set 2 (page 232)

### 1

```{r}
gss_cat %>% 
  ggplot() +
  geom_bar(aes(tvhours))
```

### 2

```{r}
sapply(select_if(gss_cat, is.factor), FUN = levels)
```

- `marital`: arbitrary  
- `race`: arbitrary  
- `rincome`: principled  
- `partyid`: principled  
- `relig`: arbitrary  
- `denom`: arbitrary  

### 3

Default value for argument `after` is `0L`.

```{r}
ggplot(rincome, aes(age, fct_relevel(rincome, "Not applicable", after = Inf))) +
  geom_point()
```

# Modifying factor levels

Use `fct_recode` to change levels in a factor.

```{r}
gss_cat %>%
  mutate(partyid = fct_recode(partyid,
                              "Republican, strong"     = "Strong republican",
                              "Republican, weak"       = "Not str republican",
                              "Independent, near rep"  = "Ind,near rep",
                              "Independent, near dem"  = "Ind,near dem",
                              "Democrat, weak"         = "Not str democrat",
                              "Democrat, strong"       = "Strong democrat"
  )) %>%
  count(partyid)
```

To combine groups, assign them to the same new level

```{r}
gss_cat %>%
  mutate(partyid = fct_recode(partyid,
                              "Republican, strong"    = "Strong republican",
                              "Republican, weak"      = "Not str republican",
                              "Independent, near rep" = "Ind,near rep",
                              "Independent, near dem" = "Ind,near dem",
                              "Democrat, weak"        = "Not str democrat",
                              "Democrat, strong"      = "Strong democrat",
                              "Other"                 = "No answer",
                              "Other"                 = "Don't know",
                              "Other"                 = "Other party"
  )) %>%
  count(partyid)
```

```{r}
# fct_collapse can be used to re-level factors too
gss_cat %>%
  mutate(partyid = fct_collapse(partyid,
                                other = c("No answer", "Don't know", "Other party"),
                                rep = c("Strong republican", "Not str republican"),
                                ind = c("Ind,near rep", "Independent", "Ind,near dem"),
                                dem = c("Not str democrat", "Strong democrat")
  )) %>%
  count(partyid)
```

```{r}
# Use fct_lump to lump together smaller levels
# it will continue aggregating while the aggregate is still the smallest group
gss_cat %>%
  mutate(relig = fct_lump(relig)) %>%
  count(relig)
# this behaviour can be over-written with the n parameter
gss_cat %>%
  mutate(relig = fct_lump(relig, n = 10)) %>%
  count(relig, sort = TRUE) %>%
  print(n = Inf)
```

## Section exercises (page 235)

### 1 - Time series of `partyid`

```{r}
gss_cat %>%
  mutate(partyid = fct_collapse(partyid,
                                other = c("No answer", "Don't know", "Other party"),
                                rep = c("Strong republican", "Not str republican"),
                                ind = c("Ind,near rep", "Independent", "Ind,near dem"),
                                dem = c("Not str democrat", "Strong democrat"))) %>%
  count(year, partyid) %>%
  group_by(year) %>%
  mutate(p = n / sum(n)) %>%
  ggplot(aes(x = year, y = p, colour = fct_reorder2(partyid, year, p))) +
  geom_point() +
  geom_line() +
  labs(x = "Year", y = "Proportion", colour = "Party ID.")

# gss_cat %>%
#   mutate(partyid = fct_collapse(partyid,
#                                 other = c("No answer", "Don't know", "Other party"),
#                                 rep = c("Strong republican", "Not str republican"),
#                                 ind = c("Ind,near rep", "Independent", "Ind,near dem"),
#                                 dem = c("Not str democrat", "Strong democrat"))) %>%
#   group_by(year, partyid) %>% 
#   tally() %>% 
#   ungroup() %>% 
#   group_by(year) %>%
#   mutate(p = n / sum(n)) %>%
#   ggplot(aes(x = year, y = p, colour = fct_reorder2(partyid, year, p))) +
#   geom_point() +
#   geom_line() +
#   labs(x = "Year", y = "Proportion", colour = "Party ID.")
```

### 2 - Collapse `rincome`

```{r}
levels(gss_cat$rincome)
new_rincome <- gss_cat %>% 
  dplyr::select(rincome) %>% 
  mutate(rincome = fct_collapse(rincome,
                                other = c("No answer", "Don't know", "Refused", "Not applicable"))) %>% 
  pull()
levels(new_rincome)
```

