---
title: "str functions"
author: "Daniel Czarnievicz"
date: "5/18/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(magrittr)
library(tidyverse)
library(stringr)
```

# str_functions

- `str_length()`: returns the lenght of a character string  
- `str_c()`: concatenates  
- `str_sub()`: subsetting strings by start and end position  
- `str_to_lower()`, `str_to_upper()`, `str_to_title()`: change capitalization of a string  
- `str_order()` and `str_sort()`: same as order and sort but with locale option
- `str_view()` and `str_view_all()`: shows matches in a viewing device  
- `str_detect()`: detect the presence or absence of a pattern in a string
- `str_locate()`: returns a matrix with start and end position of patterns
- `regex()`: greater control on regex definition
- `fixed()`: matches exactly the specified sequence of bytes
- `coll()`: same as fixed but respects human character comparison rules. compares strings using standard collation rules. Takes locale.

# Character classes

- `.` matches anything except for whitespace  
- `\d` matches any digit  
- `\s` matches any whitespace  
- `[abc]` matches a, b or c  
- `[^abc]` matches everything that is not a nor b nor c  

# Repetition

- `?`: 0 or 1 time  
- `+`: 1 or more times  
- `*`: 0 or more times  
- `{n}`: exactly n times  
- `{n,}`: n or more times  
- `{,m}`: at most m times  
- `{n,m}`: between n and m times  

# Escaping  reserved (meta-)characters

`\` is a reserved character. To create reg exps that contain a `\` you'll need to escape it.

```{r error=TRUE}
str_view(string = "This string contains the following digits: 1645454313149913",
         pattern = "\d")
```

```{r}
str_view(string = "This string contains the following digits: 1645454313149913",
         pattern = "\\d")
```

```{r}
str_view_all(string = "This string contains the following digits: 1645454313149913",
             pattern = "\\d")
```

- `"\"`: This will escape the next character in the R string.
- `"\\"`: This will resolve to `\` in the regular expression, which will escape the next character in the regular expression.
- `"\\\"`: The first two backslashes will resolve to a literal backslash in the regular expression, the third will escape the next character. So in the regular expression, this will escape some escaped character.
- `"\\\\"`: This will match a literal blackslash `\` is a character string  

```{r}
str_view(string = "This string contains one backslash \\",
         pattern = "\\\\")
```

